---
title: "Assignment 1: Task 2"
author: "Jessica Jagdeo"
date: "2/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

Environmental and climatic variables for a range of countries have been obtained from publicly available remote sensing datasets and published on the Google Earth Engine platform (zander_venter, Kaggle). This project explores the potential associations between the country-level environmental and climatic variables through a principal components analysis. The goal is to determine if there are any associations between countries' accessibility to cities, elevation, and mean annual temperature and precipitation. 

#### 1. Load necessary packages.
```{r}

library(tidyverse)
library(janitor)
library(kableExtra)
library(naniar)
library(skimr)
library(sf)
library(tmap)
library(ggfortify)

```

#### 2. Read in the data: Country-level environmental and climatic variables from zander_venter on Kaggle.
```{r}

# I am interested in associations between a country's accessiblity to cities, elevation, average annual rainfall, and average annual temperature, so I will simplify the raw data set to contain only these variables

env_var <- read_csv("world_env_vars.csv") 

env_simplified <- env_var %>% 
  clean_names() %>% 
  select(country, accessibility_to_cities, elevation, temp_mean_annual, rain_mean_annual)

```

#### 3. Simplify the data for PCA analysis.
```{r}

# Check out NAs within the simplified data frame to assess missing data points

summary(env_simplified)

gg_miss_var(env_simplified)

# There are missing data points, but not many compared to the actual scope of the data (10 out of 243 observations), so I will retain only complete observations

```

```{r}

# Using drop_na() to remove observations that contain at least 1 NA from simplified data set

env_simp_nona <- env_simplified %>%
  drop_na() %>% 
  select(-country) %>% 
  rename("Accessibility to Cities" = accessibility_to_cities, "Elevation" = elevation, "Mean Annual Temperature" = temp_mean_annual, "Mean Annual Precipitation" = rain_mean_annual)

summary(env_simp_nona) # Recheck for NAs

```

#### 4. Conduct a Principal Components Analysis.
```{r}

# Principal Components Analysis

env_var_pca <- prcomp(env_simp_nona, scale = TRUE)

summary(env_var_pca)

# Plot PCA results to visualize associations

env_var_biplot <- autoplot(env_var_pca, 
                      colour = "NA",
                      loadings.colour = "blue",
                      loadings.label = TRUE,
                      loadings.label.size = 3,
                      loadings.label.colour = "black",
                      loadings.label.repel = TRUE) +
                      theme_classic() + 
                      labs(x = "Axis 1 (43.35%)",
                           y = "Axis 2 (25.17%)")
  

env_var_biplot

# Biplot captures about 68% of variance

```

**Figure 1. Principal components analysis biplot showing the associations between environmental and climatic variables at the country-level.** Variables depicted by this biplot include accessiblity to cities, mean annual precipitation, mean annual temperature, and elevation. Data source: zander_venter, Kaggle.

To summarize key observations from Figure 1:

1) There is no correlation between accessibility to cities and elevation. 

2) There is also no correlation between accessibility to cities and mean annual temperature. 

3) There is a negative correlation between elevation and mean annual temperature, so as elevation increases, mean annual temperature decreases and vice versa. 